---
title: "survival_analysis"
format: html
editor: visual
---

# Survival Analysis with Gamma Regression

This is a small demonstration of Gamma regression. Data is not censored. The dataset and the analysis is based on the book "Analysis of Variance, Design and Regression: Linear Modeling for Unbalanced Data", Second Edition, Ronald Christensen, Chapter 22 "Exponential and Gamma Regression: Time-to-Event Data".

The assumption of the model we are going to use is that each observation in the dataset comes from a Gamma distribution. That is

$$
\begin{aligned}
Y &\sim Gamma(\alpha, \lambda), \\
f(y;\alpha, \lambda) &= \frac{\lambda^\alpha}{\Gamma(\alpha)}\exp(-\lambda y)y^{\alpha -1}
\end{aligned}
$$

For $\alpha=1$ we obtain the exponential distribution. The mean of the Gamma distribution is $\frac{\alpha}{\lambda}$ and its variance is $\frac{\alpha}{\lambda^2}$. The following linear model is used:

$$
\begin{aligned}
\log(\frac{\alpha}{\lambda_i}) &= x_i' \beta
\end{aligned}
$$

That is we assume that the log of the mean is a linear function of the covariates.

## Dataset

We load the dataset for our analysis.

```{r}
df <- read.table("data.csv", sep = "\t", header = TRUE)
dim(df)
```

We see the dataset consists of 33 rows and has three columns.

```{r}
head(df)
```

-   The y column represents: number of weeks a patient survived after diagnosis of acute myelogenous leukemia.

-   WBC stands for White Bood Cell Count at diagnosis.

-   AG is a grouping variable (taking values 1 for AG positive and 2 for AG negative)

```{r}
hist(df$y,
     main  = "Histogram of y",
     xlab  = "y",
     col   = "steelblue",
     border = "white",
     breaks = 30)
```

We see as expected no negative values in y. Also we observe that most values are clustered fairly close to zero.

```{r}
# Since AG takes only two values (AG positive and AG negative) we use it as a factor in
# the anaylsis.
df$AG <- as.factor(df$AG)
```

## Analysis

We use the glm package in R to perform a Gamma regression.

```{r}
# We use the following model hierarchy.
model1 <- glm(y ~ 1, data = df, family = Gamma(link = "log"))
model2 <- glm(y ~ AG, data = df, family = Gamma(link = "log"))
model3 <- glm(y ~ WBC + AG, data = df, family = Gamma(link = "log"))
```

**We have a hierarchy of nested models:**

-   `model1` is the baseline (intercept only)

-   `model2` tests whether AG alone explains y

-   `model3` tests whether WBC adds explanatory power on top of AG

Since the models are nested we can use an Analysis of Deviance.

```{r}
anova(model1, model2, model3, test = "LRT")
```

We stick to model2 since the p-value for the comparison of the second and third model is not significant..

```{r}
summary(model2)
```

### Model Checks

```{r}
library(statmod)
library(ggplot2)
library(patchwork)  # to combine plots

# Extract residuals and fitted values
rqr       <- qresid(model2)
dev_res   <- residuals(model2, type = "deviance")
fitted_v  <- fitted(model2)
cooks_d   <- cooks.distance(model2)
n         <- length(fitted_v)

diag_data <- data.frame(
  fitted   = fitted_v,
  rqr      = rqr,
  dev_res  = dev_res,
  cooks    = cooks_d,
  index    = 1:n
)

# 1. QQ Plot (quantile residuals)
p1 <- ggplot(diag_data, aes(sample = rqr)) +
  stat_qq(alpha = 0.6, color = "steelblue") +
  stat_qq_line(color = "coral", linewidth = 1) +
  labs(title = "QQ Plot",
       subtitle = "Randomised Quantile Residuals",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()

# 2. Residuals vs Fitted
p2 <- ggplot(diag_data, aes(x = fitted, y = rqr)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_hline(yintercept = 0, 
             linetype   = "dashed", 
             color      = "grey40") +
  geom_smooth(se      = FALSE, 
              color   = "coral", 
              method  = "loess") +
  labs(title    = "Residuals vs Fitted",
       subtitle = "Randomised Quantile Residuals",
       x        = "Fitted Values",
       y        = "Quantile Residuals") +
  theme_minimal()

# 3. Scale-Location plot
p3 <- ggplot(diag_data, aes(x = fitted, y = sqrt(abs(rqr)))) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(se     = FALSE, 
              color  = "coral", 
              method = "loess") +
  labs(title    = "Scale-Location",
       subtitle = "Check for homoscedasticity",
       x        = "Fitted Values",
       y        = expression(sqrt("|Quantile Residuals|"))) +
  theme_minimal()

# 4. Cook's Distance
p4 <- ggplot(diag_data, aes(x = index, y = cooks)) +
  geom_bar(stat  = "identity", 
           fill  = "steelblue", 
           alpha = 0.6) +
  geom_hline(yintercept = 4/n,       # common threshold
             linetype   = "dashed", 
             color      = "coral",
             linewidth  = 0.8) +
  geom_text(data = subset(diag_data, cooks > 4/n),
            aes(label = index),
            vjust = -0.5, size = 3, color = "coral") +
  labs(title    = "Cook's Distance",
       subtitle = paste0("Dashed line = 4/n threshold (", round(4/n, 3), ")"),
       x        = "Observation Index",
       y        = "Cook's Distance") +
  theme_minimal()

# Combine all four plots
(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title    = "Gamma GLM Diagnostic Plots (model2)",
    subtitle = "Randomised quantile residuals used throughout"
  )
```

### Interpretation

Coefficients are on the log scale. We can exponentiate them for easier interpretability:

```{r}
exp(coef(model3))         # Multiplicative effect on y
exp(confint(model3))      # Exponentiated confidence intervals
```

-   The coefficient for `AG2` is approx. `0.33`, meaning switching from AG1 to AG2 is associated with a **67% decrease** in `y`.

-   The coefficient for `WBC` is approx. `1.0`. Furthmore, its confidence interval contains 1.0. Therefore, the analysis does not provide any evidence that WBC has an effect on y.

This confirms that we can stick to model2. Since AG has only two levels in model 2 we can compare the distributions predicted by the model and the histograms coming from the data for the two groups.

```{r}
# We get the shape parameter directly from the model
dispersion <- summary(model2)$dispersion
shape      <- 1 / dispersion

# Define x_range
x_range <- seq(min(df$y), max(df$y), length.out = 500)

# Predicted means from the model for each AG group (at mean WBC)
mean_WBC <- mean(df$WBC)

mean_AG1 <- exp(coef(model2)["(Intercept)"])

mean_AG2 <- exp(coef(model2)["(Intercept)"] + 
                coef(model2)["AG2"])

# Gamma rate = shape / mean
rate_AG1 <- shape / mean_AG1
rate_AG2 <- shape / mean_AG2

# -------------------------------------------------------
# Build PDF curves instead of CDF
pdf_data <- rbind(
  data.frame(y   = x_range,
             pdf = dgamma(x_range, shape = shape, rate = rate_AG1),
             AG  = "AG1"),
  data.frame(y   = x_range,
             pdf = dgamma(x_range, shape = shape, rate = rate_AG2),
             AG  = "AG2")
)

ggplot() +
  geom_histogram(data = subset(df, AG == "1"),
                 aes(x = y, y = after_stat(density), fill = "AG1"),
                 alpha = 0.4, bins = 30, color = "white") +
  geom_histogram(data = subset(df, AG == "2"),
                 aes(x = y, y = after_stat(density), fill = "AG2"),
                 alpha = 0.4, bins = 30, color = "white") +
  # PDF lines — same scale as density histogram
  geom_line(data = pdf_data,
            aes(x = y, y = pdf, color = AG), linewidth = 1) +
  scale_fill_manual(values  = c("AG1" = "steelblue", "AG2" = "coral")) +
  scale_color_manual(values = c("AG1" = "steelblue", "AG2" = "coral")) +
  labs(title    = "Histogram and Model-Based PDF for AG1 and AG2",
       subtitle = "PDFs derived from Gamma GLM (model2)",
       x = "y", y = "Density",
       fill = "Group", color = "Group") +
  theme_minimal()
```

Let us check for example what is the probability of surviving more than 7 weeks in the two groups the model predicts:

```{r}
dispersion <- summary(model2)$dispersion
shape      <- 1 / dispersion

# Predicted means from model2
mean_AG1 <- exp(coef(model2)["(Intercept)"])
mean_AG2 <- exp(coef(model2)["(Intercept)"] + coef(model2)["AG2"])

# Gamma rates
rate_AG1 <- shape / mean_AG1
rate_AG2 <- shape / mean_AG2

# Survival probability more than 7 weeks
surv_AG1 <- 1 - pgamma(7, shape = shape, rate = rate_AG1)
surv_AG2 <- 1 - pgamma(7, shape = shape, rate = rate_AG2)

cat("P(survive > 7 weeks | AG1):", round(surv_AG1, 4), "\n")
cat("P(survive > 7 weeks | AG2):", round(surv_AG2, 4), "\n")
```

Let us look also at the survival curves.

```{r}
library(ggplot2)

dispersion <- summary(model2)$dispersion
shape      <- 1 / dispersion

# Predicted means from model2
mean_AG1 <- exp(coef(model2)["(Intercept)"])
mean_AG2 <- exp(coef(model2)["(Intercept)"] + coef(model2)["AG2"])

# Gamma rates
rate_AG1 <- shape / mean_AG1
rate_AG2 <- shape / mean_AG2

# Survival probabilities at 7 weeks
surv_AG1 <- 1 - pgamma(7, shape = shape, rate = rate_AG1)
surv_AG2 <- 1 - pgamma(7, shape = shape, rate = rate_AG2)

# Survival curves
t_range <- seq(0, max(df$y), length.out = 500)

surv_data <- rbind(
  data.frame(t    = t_range,
             surv = 1 - pgamma(t_range, shape = shape, rate = rate_AG1),
             AG   = "AG1"),
  data.frame(t    = t_range,
             surv = 1 - pgamma(t_range, shape = shape, rate = rate_AG2),
             AG   = "AG2")
)

# 7-week annotation points
points_7 <- data.frame(
  t    = c(7, 7),
  surv = c(surv_AG1, surv_AG2),
  AG   = c("AG1", "AG2")
)

# Plot
ggplot(surv_data, aes(x = t, y = surv, color = AG)) +
  geom_line(linewidth = 1) +
  # Vertical dashed line at 7 weeks
  geom_vline(xintercept = 7, 
             linetype   = "dashed", 
             color      = "grey40") +
  # Horizontal dashed lines to y-axis
  geom_segment(data = points_7,
               aes(x = 0, xend = 7, y = surv, yend = surv, color = AG),
               linetype = "dashed", linewidth = 0.5) +
  # Points at 7 weeks
  geom_point(data  = points_7, 
             size  = 3) +
  # Annotate probabilities
  geom_text(data  = points_7,
            aes(label = paste0("P=", round(surv, 3))),
            hjust = -0.2, vjust = -0.5, size = 3.5) +
  scale_color_manual(values = c("AG1" = "steelblue", "AG2" = "coral")) +
  scale_y_continuous(limits = c(0, 1), 
                     breaks = seq(0, 1, by = 0.1),
                     labels = scales::percent) +
  scale_x_continuous(breaks = seq(0, max(df$y), by = 5)) +
  labs(title    = "Model-Based Survival Curves by AG Group",
       subtitle = "Derived from Gamma GLM (model2) | Dashed lines indicate 7-week survival probability",
       x        = "Weeks since diagnosis",
       y        = "Survival probability S(t)",
       color    = "Group") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```
